# Using ggpaintr from an LLM {#sec-llm}

Large language models are fluent at writing raw `shiny::fluidPage()` + server scaffolding from memory. They are *not* fluent at writing ggpaintr, because every model today has zero prior exposure to this package. Without help, a user asking "build me a Shiny app where I can pick columns from iris and see a scatter plot" gets hundreds of lines of hand-rolled Shiny that the `ptr_app()` one-liner would have covered.

This chapter shows how to make any [ellmer](https://ellmer.tidyverse.org/) chat session reach for ggpaintr first, using two exported helpers:

- `ptr_llm_primer()` — the system-prompt preamble that teaches the model the three-level integration model (@sec-use-cases) and the "call the docs tool before writing code" rule.
- `ptr_llm_register()` — one-line registration of the `ggpaintr_docs` tool that serves focused runnable examples on demand.

The tool is backed by `ptr_llm_topic()`, which returns the content of one bundled topic file. The LLM pulls only the topic it needs, instead of the whole manual loading the context window on every turn.

## Quick start — three lines of setup

```{r}
#| eval: false
library(ellmer)
library(ggpaintr)

chat <- chat_anthropic(system_prompt = ptr_llm_primer())
ptr_llm_register(chat)

chat$chat("Build a Shiny app where the user picks X and Y columns from mtcars.")
```

Under the hood:

- `chat_anthropic()` loads the primer (~4,000 tokens) into the system prompt, which stays in context every turn.
- `ptr_llm_register()` attaches a single tool named `ggpaintr_docs` to the chat session. The tool's one argument — `topic` — is constrained by `type_enum()` to the exact set of topic names returned by `ptr_llm_topics()`, so the LLM cannot request a topic that does not exist.

On the first turn the model sees the user's prompt, recognises it as a plot-explorer task, calls `ggpaintr_docs(topic = "level1_ptr_app")`, and writes a `ptr_app(...)` snippet rather than raw Shiny.

## What the primer + tool split buys you

A common alternative is to stuff the entire manual into the system prompt. That works but wastes tokens: every chat turn pays for the full manual whether or not it is relevant. The primer/tool split is a lightweight form of retrieval:

| Layer                  | When it loads              | What it teaches |
|------------------------|----------------------------|------------------|
| Primer                 | Every turn (~4,000 tokens) | ggpaintr exists; 3-level model; decision rule. |
| `ggpaintr_docs(topic)` | On demand                  | One focused runnable example (~500–2,800 tokens). |

The primer carries an explicit rule — *"Before writing R code for an interactive plot task, call `ggpaintr_docs(topic)` to fetch the runnable example"* — and that rule plus the enum-constrained argument is what makes the LLM reliably reach for the tool instead of hallucinating an API.

## The available topics

This chunk runs against the pinned ggpaintr at render time, so the list below is the canonical one:

```{r}
#| eval: true
library(ggpaintr)
ptr_llm_topics()
```

A short guide to each:

| Topic                  | When the LLM should call it |
|------------------------|------------------------------|
| `overview`             | First orientation — the 3-level model and decision rule. |
| `formula_syntax`       | Which placeholder keyword goes where; pipelines; `shared = "..."`. |
| `level1_ptr_app`       | Quick turn-key single-plot app (`ptr_app()`). |
| `level1_ptr_options`   | Package-global settings: verbosity, layer-checkbox default. |
| `level2_module`        | Embed the default-layout block (`ptr_ui()` / `ptr_server()`) in the user's own Shiny app. |
| `level2_custom_ids`    | Id collisions, the `ptr_` reserved prefix, generated-id grammar. |
| `level2_shared`        | Multiple instances + the shared-coordinator trio for cross-instance widgets. |
| `level2_ui_text`       | Override labels, help, placeholders; multi-section cascade. |
| `level3_custom_render` | Own `renderPlot()` / `renderPlotly()` reading `state$runtime()`. |
| `level3_layout`        | Own the layout — bare pieces, the two combinators, the optional `ptr_ui_page()` shell. |
| `level3_gg_extra`      | Round-trip host layers into the generated code pane. |
| `custom_placeholder`   | Date pickers, sliders, any widget not in the five built-ins. |

Topics whose example is a standalone runnable app open with a `library()` preamble; fragment-style topics (`overview`, `level2_custom_ids`, `level3_custom_render`, `level3_layout`) do not, by design.

## Inspecting what the LLM will see

You can read any topic directly, without booting a chat session — useful for integration tests or checking content before release. These run at render time:

```{r}
#| eval: true
cat(substr(ptr_llm_primer(), 1, 400), "\n...")
```

```{r}
#| eval: true
cat(substr(ptr_llm_topic("level1_ptr_app"), 1, 600), "\n...")
```

## Provider swap and model choice

The tool definition is provider-neutral — the same registration line works with every chat constructor ellmer exports:

```{r}
#| eval: false
# Anthropic
chat <- chat_anthropic(system_prompt = ptr_llm_primer(),
                       model = "claude-sonnet-4-6")
ptr_llm_register(chat)

# OpenAI
chat <- chat_openai(system_prompt = ptr_llm_primer(),
                    model = "gpt-4o-mini")
ptr_llm_register(chat)

# Google
chat <- chat_google_gemini(system_prompt = ptr_llm_primer())
ptr_llm_register(chat)

# Local (ollama)
chat <- chat_ollama(system_prompt = ptr_llm_primer(),
                    model = "llama3.1")
ptr_llm_register(chat)
```

The primer is small; any model that supports tool calling will work. For fast iteration on the registration itself, a small, cheap model is a reasonable default.

## Manual registration — what the helper does

`ptr_llm_register()` is a thin convenience wrapper. To customise the tool (different name, tighter description, extra arguments), do it by hand:

```{r}
#| eval: false
library(ellmer)

ggpaintr_docs <- tool(
  ptr_llm_topic,
  name        = "ggpaintr_docs",
  description = paste(
    "Fetch a runnable ggpaintr example for one integration topic.",
    "Call before writing R code for interactive ggplot tasks."
  ),
  arguments = list(
    topic = type_enum(
      ptr_llm_topics(),
      "Which topic to fetch."
    )
  )
)

chat <- chat_anthropic(system_prompt = ptr_llm_primer())
chat$register_tool(ggpaintr_docs)
```

The pieces: `tool(...)` is ellmer's constructor wrapping the exported R function; `type_enum(ptr_llm_topics(), ...)` constrains the LLM's topic choice to the exact names the package ships; `chat$register_tool(...)` attaches it to the session. If you use this form, refresh the enum whenever the package updates — the helper snapshots the list once at registration time.

## Testing without spending tokens

The ellmer `tool()` object wraps the underlying R function — call it directly for tests or sanity checks, no chat session and no API key involved. This runs at render time:

```{r}
#| eval: true
payload <- ptr_llm_topic("level2_module")
nchar(payload)
```

`ptr_llm_topics()` returns the canonical list; if you add a topic in a downstream repo, assert the name is present before shipping the registration.

## When to re-register

- **After upgrading ggpaintr** — new topics are not reachable until `ptr_llm_register()` runs again (the enum list is snapshotted).
- **Switching chat providers** — registration is per-`Chat` object; every new session needs its own registration.
- **Changing the tool name** — drop the old tool with `chat$set_tools()` (inspect what is attached via `chat$get_tools()`) and register the new one; ellmer has no dedicated `unregister_tool()`, but you do not need to rebuild the chat.

## Current behavior boundary

- The primer and topic files are bundled with the package. Their factual blocks — signatures, the token table, the topic index — are regenerated from live introspection by the package's `dev/build-llm.R` (lint / generate / verify modes); the surrounding prose is still hand-curated.
- ellmer is a suggested dependency — nothing loads until you call the helper, and `ptr_llm_register()` errors with a clear message if ellmer is not installed.
- The helper sets only the tool. The system prompt is your responsibility — pass `ptr_llm_primer()` to the `chat_*()` constructor explicitly.
